---
title: "Statistical Age–Length Key (ALK) Modeling Framework"
author: "AFSC Assessment Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Age–Length Key (ALK) Modeling Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette describes the statistical modeling framework used to construct age–length keys (ALKs) within this package. The approach replaces traditional empirical ALKs with a hierarchical generalized additive modeling (GAM) structure that:

- Smooths the age–length relationship
- Allows borrowing of information across years and areas within region
- Maintains independence among major geographic regions
- Produces posterior age distributions for each length observation

The resulting age compositions are suitable for use in Stock Synthesis.

# Regional Independence

Models are fit **independently for each `REGION_GRP`**.

No parameters are shared across regions. This ensures:

- Biological differences among regions are preserved
- Borrowing occurs only within region
- No cross-region smoothing or shrinkage occurs

# Modeling Structure

The ALK framework proceeds in two stages:

1. Survey backbone model  
2. Fishery quarter deviation model  

## 1. Survey Backbone Model

Survey age data (Q3 backbone) define the baseline age–length relationship.

Age is modeled as:

$$
A_i \sim \mathcal{N}(\mu_i, \sigma^2)
$$

$$
\mu_i = f(L_i) + b_{YEAR(i)} + b_{AREA(i)}
$$

Where:

- $A_i$ = observed age  
- $L_i$ = length  
- $f(L)$ = penalized spline smooth  
- $b_{YEAR}$ = random year effect  
- $b_{AREA}$ = random area effect  

### Interpretation

- The spline produces a smooth age–length curve.
- Random effects allow partial pooling across years and areas within region.
- Sparse years/areas shrink toward the regional mean.
- If predicting an unseen year or area, the random effect is set to zero.

## 2. Fishery Quarter Deviation Model

Fishery ages are modeled as deviations from the survey backbone.

$$
\Delta_i = A_i - \hat{\mu}^{survey}_i
$$

$$
\Delta_i = g(L_i, Q_i) + u_{YEAR(i)} + u_{AREA(i)}
$$

Where:

- $g(L,Q)$ = smooth interaction between length and quarter
- $u_{YEAR}, u_{AREA}$ = random effects

Predicted age becomes:

$$
\hat{A}_i = \hat{\mu}^{survey}_i + \hat{\Delta}_i
$$

This structure separates:

- Baseline biological growth
- Seasonal and fishery-specific deviations

# Posterior Age Assignment

For each length observation, a posterior age distribution is computed:

$$
P(A=a \mid L,s) \propto 
\phi(a \mid \hat{\mu}(L,s), \hat{\sigma})
\times
\pi(a \mid s)
$$

Where:

- $\phi$ = model likelihood (normal density)
- $\pi(a \mid s)$ = empirical sex-specific prior
- $s$ = sex

Posterior probabilities are normalized over ages:

$$
a = 0, 1, \dots, A_{max}
$$

Unsexed fish are split 50/50 between male and female posteriors.

# Borrowing Structure

Borrowing is controlled explicitly:

| Dimension | Borrowing | Mechanism |
|------------|-----------|------------|
| Across regions | No | Separate models |
| Across years (within region) | Yes | Random effects |
| Across areas (within region) | Yes | Random effects |
| Across quarters | Partial | Smooth interaction |

If a year or area is not observed in training data:

- The corresponding random effect is set to zero
- Prediction reverts to the regional smooth

# Computational Implementation

Models are fit using `mgcv`.

For large datasets, `bam()` may be used for efficiency.

Posterior computations:

1. Are performed on unique `(REGION_GRP, YEAR, QUARTER, AREA_K, SEX2, LENGTH)` combinations  
2. Use vectorized prediction for efficiency  
3. Are merged back to full length-frequency data  

# Example Workflow

```{r example-workflow, eval=FALSE}
# Fit age predictor
age_pred <- fit_age_predictor(
  survey_age_dt  = survey_data,
  fishery_age_dt = fishery_data,
  region_col     = "REGION_GRP"
)

# Predict age distributions from length data
lf_with_age <- predict_age_from_lf(
  lf_dt     = fishery_lengths,
  predictor = age_pred,
  max_age   = 10
)

# Construct catch-weighted age compositions
age_comps <- LENGTH_AGE_BY_CATCH(
  lf_dt = lf_with_age,
  ...
)
```

# Advantages Over Empirical ALKs

Compared to traditional empirical ALKs:

- Smooth age–length curve reduces sampling noise
- Partial pooling stabilizes sparse strata
- Explicit seasonal structure
- Region-specific independence
- Full posterior age distributions

# Use in Stock Synthesis

The resulting age compositions:

- Are catch-weighted
- Reflect seasonal and spatial effects
- Incorporate modeled age uncertainty
- Are formatted for SS input files

# Summary

This modeling framework provides a statistically robust, region-specific, and computationally efficient approach to constructing age–length keys for assessment applications.

It balances smoothing and biological realism while maintaining explicit control over spatial borrowing.
